---
title: "Final Project"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
---

```{r setup, include=FALSE}
require(flexdashboard)
require(ggplot2)
require(plotly)
library(tidyverse)
library(gridExtra)
library(plotly)
library(cowplot)
library(tigris)
library(gghighlight)
library(statebins)
library(MetBrewer)
library(here)
```

```{r}

#Load data
math_achievement <- rio::import("~/Documents/EDLD-652/finalproject/Data/Eighth-grade math achievement levels.xlsx")
science_achievement_1 <-  rio::import("~/Documents/EDLD-652/finalproject/Data/Eighth-grade science achievement levels.xlsx")
science_achievement_2 <-  rio::import("~/Documents/EDLD-652/finalproject/Data/Eighth-grade science achievement levels (1).xlsx")
funding <-  rio::import("~/Documents/EDLD-652/finalproject/Data/Per-pupil educational expenditures.xlsx")


```


```{r}
#Rearrange data

# Math 
#Rename column and covert data to numeric
math_achievement <- math_achievement %>% 
  rename("AchievementLevel" = "Achievement Level") %>% 
  mutate(Data = as.numeric(Data), TimeFrame = as.numeric(TimeFrame)) %>% 
   mutate(AchievementType = case_when(grepl("basic", AchievementLevel) ~ "Basic",
                            grepl("proficient", AchievementLevel, ignore.case = TRUE) ~"Proficient"),
          Subject = "Math") 
# Science
#rbind both science achievement files to create one with all years, rename column, and convert data to numeric
science_achievement <- rbind(science_achievement_1, science_achievement_2) %>% 
  rename("AchievementLevel" = "Achievement Level") %>% 
  mutate(Data = as.numeric(Data), TimeFrame = as.numeric(TimeFrame)) %>% 
   mutate(AchievementType = case_when(grepl("basic", AchievementLevel) ~ "Basic",
                            grepl("proficient", AchievementLevel, ignore.case = TRUE) ~"Proficient"),
          Subject = "Science") 
# Funding
#Convert numbers from character to numeric
funding$Data <- as.numeric(funding$Data)


#Make dataframe w/ regions
#census$State = achievement$Location (not US)
```


```{r}
#Set defaults
theme_set(theme_minimal())

options(ggrepel.max.overlaps = 25)
```


Sidebar Title {.sidebar}
==========================
Your text here. You can use markdown syntax, including 
[links](http://blah.com), *italics*, **bolding**, etc.

The data used for this project come from the KIDS COUNT Study, which is conducted by the Annie E. Casey Foundation (https://datacenter.kidscount.org/). Data is collected at both the national level and the state level within the US. The KIDS COUNT project contains data related to a number of different aspects of child development and well-being, including education, health, risky behavior, and family/community relationships. 

For the specific research questions I’m interested in looking at, achievement data are available for fourth and eighth graders. Data were collected every two years, from 2003-2022 (data were released in 2022 instead of 2021). Data are available per state, and nationally. Achievement data are represented using 4 categories (below basic, at or above basic, below proficient, at or above proficient). The proficient category builds on the basic category, so below basic and at or above basic add up to 100% of students, and below proficient and at or above proficient add up to 100% of students. 

Question 1
===================================== 

How has educational achievement in math and science changed over time?

Row {data-height=400}
-----------------------------------------

### Chart A 

This is **typical** .Rmd text.

```{r}
science_plot <- science_achievement %>% 
  filter(Location=="United States") %>% 
  ggplot(aes(x=TimeFrame, y=Data, group = AchievementLevel, color=AchievementLevel)) +
   geom_point() +
  geom_line() +
  scale_color_manual(values=met.brewer("Morgenstern", 4)) +
  labs(x="School Year", y = "Proportion", title = "Science Achievement (US average)", color = "Achievement Level")

math_plot <- math_achievement %>% 
  filter(Location=="United States") %>% 
  ggplot(aes(x=TimeFrame, y=Data, group = AchievementLevel, color=AchievementLevel)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values=met.brewer("Morgenstern", 4)) +
  labs(x="School Year", y = "Proportion", title = "Math Achievement (US average)", color = "Achievement Level")+
  theme(axis.text.x = element_text( angle = 45))

grid.arrange(science_plot, math_plot)
```



### Chart B {data-padding=20}

```{r}

subject_achievement_plot <- ggplot() + 
geom_line(data=science_achievement %>% 
  filter(Location=="United States", AchievementLevel == "At or above proficient"| AchievementLevel=="At or above basic"), aes(x=TimeFrame, y=Data, group = AchievementLevel, color=Subject, linetype=AchievementLevel)) + 
   geom_line(data=math_achievement %>% 
  filter(Location=="United States", AchievementLevel == "At or above proficient"| AchievementLevel=="At or above basic"), aes(x=TimeFrame, y=Data, group = AchievementLevel, color=Subject, linetype=AchievementLevel)) + 
  scale_color_manual(values=met.brewer("Renoir", 2)) +
  labs(x="School Year", y = "Proportion", title = "8th Grade Science & Math Achievement (US average)", linetype= "Achievement Level") 

subject_achievement_plot
```

   
Question 2
=====================================   

How does educational achievement in math and science differ across states?

```{r}
#Maps for achievement
st <- states()

# Math, 2019
math_2019 <- math_achievement %>% 
         filter(LocationType=="State" & TimeFrame==2019 & (AchievementLevel == "At or above basic" | AchievementLevel=="At or above proficient")) %>% 
  rename(NAME = Location) 

math_map <- left_join(st, math_2019)

math_map <- shift_geometry(math_map,position = 'below')

math_map_filtered <- math_map %>% 
  filter(!is.na(AchievementLevel))

#Science, 2015
science_2015 <- science_achievement %>% 
         filter(LocationType=="State" & TimeFrame==2015 & (AchievementLevel == "At or above basic" | AchievementLevel=="At or above proficient")) %>% 
  rename(NAME = Location) 

science_map <- left_join(st, science_2015)

science_map <- shift_geometry(science_map,position = 'below')

science_map_filtered <- science_map %>% 
  filter(!is.na(AchievementLevel))
```

Row {data-height=400}
-----------------------------------------

### Chart C1
    
```{r}
#Maps for achievement
#Math, 2019
ggplot(math_map_filtered) + 
geom_sf(aes(fill = Data)) +
    facet_wrap(~AchievementLevel) +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title = "Math Achievement (2019)", caption="Data from most recent year available") +
  theme_void()
```

### Chart C2
```{r}
# Science

ggplot(science_map_filtered) + 
geom_sf(aes(fill = Data)) +
  facet_wrap(~AchievementLevel) +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title = "Science Achievement (2015)", caption="Data from most recent year available") +
  theme_void()
```
  
Row {data-height=400}
-----------------------------------------
  
### Chart D

```{r}
statebins(science_2015 %>% filter(AchievementLevel == "At or above basic"),
state_col = "NAME",
value_col = "Data") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Science Achievement, Basic (2019)") +
  theme_statebins("right")
```

### Chart E

```{r}
statebins(science_2015 %>% filter(AchievementLevel == "At or above proficient"),
state_col = "NAME",
value_col = "Data") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Science Achievement, Proficient (2019)") +
  theme_statebins("right")
```

Row 
-----------------------------------------

### Chart F
```{r}
ggplot(science_2015, aes(state=NAME, fill=Data)) +
  geom_statebins() +
  coord_equal() +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  facet_wrap(~AchievementLevel) +
  labs(title="Science Achievement (2015)") +
  theme_statebins("right")
```

Row {data-height=400}
-----------------------------------------
### Chart G
```{r}
statebins(math_2019 %>% filter(AchievementLevel == "At or above basic"),
state_col = "NAME",
value_col = "Data") +
    scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Math Achievement, Basic (2019)") +
  theme_statebins("right")
```

### Chart H
```{r}
statebins(math_2019 %>% filter(AchievementLevel == "At or above proficient"),
state_col = "NAME",
value_col = "Data") +
    scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Math Achievement, Proficient (2019)") +
  theme_statebins("right")

```

Row 
-----------------------------------------

### Chart I
```{r}
ggplot(math_2019, aes(state=NAME, fill=Data)) +
  geom_statebins() +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  facet_wrap(~AchievementLevel) +
  labs(title="Math Achievement (2019)") +
  theme_statebins("right")

```

Question 3
=====================================     

a. How has funding for education changed over time? 

b. How does funding differ between states?

c. How have changes in funding differed between states?

```{r} 
#, fig.width=4, fig.height=2.5
funding_plot <- funding %>% 
  filter(Location=="United States") %>% 
  ggplot(aes(x=TimeFrame, y=Data)) +
  geom_line(aes(group=1)) +
  labs(x="Year", y = "Dollars", title = "Per-student spending (US average)", caption="Not adjusted for inflation")+
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  theme(axis.text.x = element_text(size=6, angle = 45))
```

```{r}
# filter for just states
filtered_funding <- funding %>%  
  filter(LocationType=="State") %>% 
# average across years
  select(Location, TimeFrame, Data) %>% 
  group_by(Location) %>% 
  summarise(avg_funding = mean(Data, na.rm=T)) %>%  #, across(Region)
  ungroup() %>% 
  mutate(observation = 1:n())


# ----- This section prepare a dataframe for labels ---- #
# Get the name and the y position of each label
label_data <- filtered_funding
 
# calculate the ANGLE of the labels
number_of_bar <- nrow(label_data)
angle <-  90 - 360 * (label_data$observation-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
 
# calculate the alignment of labels: right or left
# If I am on the left part of the plot, my labels have currently an angle < -90
label_data$hjust<-ifelse( angle < -90, 1, 0)
 
# flip angle BY to make them readable
label_data$angle<-ifelse(angle < -90, angle+180, angle)
# ----- ------------------------------------------- ---- #
```


Row {data-height=400}
-----------------------------------------

### Chart J
    
```{r}
#Plot

# Start the plot
p <- ggplot(filtered_funding, aes(x=as.factor(Location), y=avg_funding)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  
  # This add the bars with a blue color
  geom_bar(stat="identity", fill=alpha("skyblue", 0.7)) +
  
  # Limits of the plot = very important. The negative value controls the size of the inner circle, the positive one is useful to add size over each bar
  ylim(-100,18000) +
  
   labs(title = "Per-student expenditures", subtitle="Averaged over 1999-2019") +   #

  
  # Custom the theme: no axis title and no cartesian grid
  theme_minimal() +
  theme(
    #plot.title = element_text(hjust = 0.5, unit(c(2.1, 0, -1, 0), "cm")),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(.5,4), "cm")      # Adjust the margin to make in sort labels are not truncated!
  ) +
  
  # This makes the coordinate polar instead of cartesian.
  coord_polar(start = 0) +
  
  # Add the labels, using the label_data dataframe that we have created before
  geom_text(data=label_data, aes(x=observation, y=avg_funding+100, label=Location, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE )
 
p
```
  
### Chart K

```{r}
#Plot ordered by values
filtered_funding_ordered <- filtered_funding %>% 
  arrange(avg_funding) %>% 
  mutate(observation_ordered = 1:n())
# ----- This section prepare a dataframe for labels ---- #
# Get the name and the y position of each label
label_data_ordered <- filtered_funding_ordered
 
# calculate the ANGLE of the labels
number_of_bar_ordered <- nrow(label_data_ordered)
angle_ordered <-  90 - 360 * (label_data_ordered$observation_ordered-0.5) /number_of_bar_ordered     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
 
# calculate the alignment of labels: right or left
# If I am on the left part of the plot, my labels have currently an angle < -90
label_data_ordered$hjust<-ifelse( angle_ordered < -90, 1, 0)
 
# flip angle BY to make them readable
label_data_ordered$angle_ordered<-ifelse(angle_ordered < -90, angle_ordered+180, angle_ordered)
# ----- ------------------------------------------- ---- #


# Start the plot

p_ordered <- ggplot(filtered_funding_ordered, aes(x=as.factor(observation_ordered), y=avg_funding)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  
  # This add the bars with a blue color
  geom_bar(stat="identity", fill=alpha("skyblue", 0.7)) +
  
  # Limits of the plot = very important. The negative value controls the size of the inner circle, the positive one is useful to add size over each bar
  ylim(-100,19000) +
  
  labs(title = "Per-student expenditures", subtitle="Averaged over 1999-2019") +  
  
  # Custom the theme: no axis title and no cartesian grid
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(.25,4), "cm")      # Adjust the margin to make in sort labels are not truncated!
  ) +
  
  # This makes the coordinate polar instead of cartesian.
  coord_polar(start = 0) +
  
  # Add the labels, using the label_data dataframe that we have created before
  geom_text(data=label_data_ordered, aes(x=observation_ordered, y=avg_funding+100, label=Location, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data_ordered$angle_ordered, inherit.aes = FALSE ) 
 
p_ordered

```

Row {data-height=400}
-----------------------------------------

### Chart L

```{r}
#Line Plot for state funding

funding %>% 
  filter(LocationType=="State") %>% 
  ggplot(aes(x=TimeFrame, y=Data, group = Location, color=Location)) +
  geom_line() +
    gghighlight(max(Data) > 20000|(max(Data) < 9250)) +
  scale_color_manual(values=met.brewer("Hiroshige", 10)) + #Derain
  labs(x="Year", y = "Dollars", title = "Per-student spending (US average)", caption="Not adjusted for inflation") +
  theme(axis.text.x = element_text(size=6, angle = 45, vjust = 1, hjust=1))
```

Row {data-height=400}
-----------------------------------------

### Chart M
```{r}
#Map with averaged funding
renamed_filtered_funding <- filtered_funding %>% 
  rename(NAME = Location) 

funding_map <- left_join(st, renamed_filtered_funding)

funding_map <- shift_geometry(funding_map,position = 'below')

funding_map_filtered <- funding_map %>% 
  filter(!is.na(avg_funding))

ggplot(funding_map_filtered) + 
geom_sf(aes(fill = avg_funding)) +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title = "Per-student expenditures", subtitle = "Averaged across 1999-2019") +
  theme_void()
```

### Chart N
```{r}
statebins(filtered_funding,
state_col = "Location",
value_col = "avg_funding") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Per-student expenditures", subtitle="Averaged over 1999-2019") +
  theme_statebins("right")
```

Question 4
=====================================     

How does funding track with educational achievement in math and science?
 
Row {data-height=400}
-----------------------------------------

### Chart J

```{r}
# spending on top
# basic performance on both math and science underneath
plot_grid(funding_plot, subject_achievement_plot, ncol=1, rel_widths = c(1, 1.5))

```


