---
title: "Project Draft"
author: "Lucy Whitmore"
date: "2/27/2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(gridExtra)
library(plotly)
library(cowplot)
library(tigris)
library(gghighlight)
library(statebins)
 library(MetBrewer)
```



```{r}
#Load data
math_achievement <- rio::import(here::here("finalproject/Data", "Eighth-grade math achievement levels.xlsx"))
science_achievement_1 <- rio::import(here::here("finalproject/Data", "Eighth-grade science achievement levels.xlsx"))
science_achievement_2 <- rio::import(here::here("finalproject/Data", "Eighth-grade science achievement levels (1).xlsx"))
funding <- rio::import(here::here("finalproject/Data", "Per-pupil educational expenditures.xlsx"))

#census <- read.csv("./us census bureau regions and divisions.csv",header=TRUE)
```


```{r}
#Rearrange data

# Math 
#Rename column and covert data to numeric
math_achievement <- math_achievement %>% 
  rename("AchievementLevel" = "Achievement Level") %>% 
  mutate(Data = as.numeric(Data), TimeFrame = as.numeric(TimeFrame)) %>% 
   mutate(AchievementType = case_when(grepl("basic", AchievementLevel) ~ "Basic",
                            grepl("proficient", AchievementLevel, ignore.case = TRUE) ~"Proficient"),
          Subject = "Math") 
# Science
#rbind both science achievement files to create one with all years, rename column, and convert data to numeric
science_achievement <- rbind(science_achievement_1, science_achievement_2) %>% 
  rename("AchievementLevel" = "Achievement Level") %>% 
  mutate(Data = as.numeric(Data), TimeFrame = as.numeric(TimeFrame)) %>% 
   mutate(AchievementType = case_when(grepl("basic", AchievementLevel) ~ "Basic",
                            grepl("proficient", AchievementLevel, ignore.case = TRUE) ~"Proficient"),
          Subject = "Science") 
# Funding
#Convert numbers from character to numeric
funding$Data <- as.numeric(funding$Data)


#Make dataframe w/ regions
#census$State = achievement$Location (not US)
```


```{r}
#Set defaults
theme_set(theme_minimal())

options(ggrepel.max.overlaps = 25)
```


## Data & Source

The data used for this project come from the KIDS COUNT Study, which is conducted by the Annie E. Casey Foundation (https://datacenter.kidscount.org/). Data is collected at both the national level and the state level within the US. The KIDS COUNT project contains data related to a number of different aspects of child development and well-being, including education, health, risky behavior, and family/community relationships. 

For the specific research questions I’m interested in looking at, achievement data are available for fourth and eighth graders. Data were collected every two years, from 2003-2022 (data were released in 2022 instead of 2021). Data are available per state, and nationally. Achievement data are represented using 4 categories (below basic, at or above basic, below proficient, at or above proficient). The proficient category builds on the basic category, so below basic and at or above basic add up to 100% of students, and below proficient and at or above proficient add up to 100% of students. 

The data are available through a data portal, located at https://datacenter.kidscount.org/. The data are openly available, and don’t require registration to be accessed. 

## Research Questions

**Question 1** 
How has educational achievement in math and science changed over time?

```{r}
#, fig.width=4, fig.height=4

science_plot <- science_achievement %>% 
  filter(Location=="United States") %>% 
  ggplot(aes(x=TimeFrame, y=Data, group = AchievementLevel, color=AchievementLevel)) +
   geom_point() +
  geom_line() +
  scale_color_manual(values=met.brewer("Morgenstern", 4)) +
  labs(x="School Year", y = "Proportion", title = "Science Achievement (US average)", color = "Achievement Level")
```

```{r}
math_plot <- math_achievement %>% 
  filter(Location=="United States") %>% 
  ggplot(aes(x=TimeFrame, y=Data, group = AchievementLevel, color=AchievementLevel)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values=met.brewer("Morgenstern", 4)) +
  labs(x="School Year", y = "Proportion", title = "Math Achievement (US average)", color = "Achievement Level")+
  theme(axis.text.x = element_text( angle = 45))

grid.arrange(science_plot, math_plot)

```



```{r}
#Achievement, above basic and above proficient 
#Both
subject_achievement_plot <- ggplot() + 
geom_line(data=science_achievement %>% 
  filter(Location=="United States", AchievementLevel == "At or above proficient"| AchievementLevel=="At or above basic"), aes(x=TimeFrame, y=Data, group = AchievementLevel, color=Subject, linetype=AchievementLevel)) + 
   geom_line(data=math_achievement %>% 
  filter(Location=="United States", AchievementLevel == "At or above proficient"| AchievementLevel=="At or above basic"), aes(x=TimeFrame, y=Data, group = AchievementLevel, color=Subject, linetype=AchievementLevel)) + 
  scale_color_manual(values=met.brewer("Renoir", 2)) +
  labs(x="School Year", y = "Proportion", title = "8th Grade Science & Math Achievement (US average)", linetype= "Achievement Level") 

subject_achievement_plot
```

**Question 2**
How does educational achievement in math and science differ across states?

```{r}
#Maps for achievement
st <- states()

# Math, 2019
math_2019 <- math_achievement %>% 
         filter(LocationType=="State" & TimeFrame==2019 & (AchievementLevel == "At or above basic" | AchievementLevel=="At or above proficient")) %>% 
  rename(NAME = Location) 

math_map <- left_join(st, math_2019)

math_map <- shift_geometry(math_map,position = 'below')

math_map_filtered <- math_map %>% 
  filter(!is.na(AchievementLevel))

ggplot(math_map_filtered) + 
geom_sf(aes(fill = Data)) +
    facet_wrap(~AchievementLevel) +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title = "Math Achievement (2019)", caption="Data from most recent year available") +
  theme_void()


# Science
science_2015 <- science_achievement %>% 
         filter(LocationType=="State" & TimeFrame==2015 & (AchievementLevel == "At or above basic" | AchievementLevel=="At or above proficient")) %>% 
  rename(NAME = Location) 

science_map <- left_join(st, science_2015)

science_map <- shift_geometry(science_map,position = 'below')

science_map_filtered <- science_map %>% 
  filter(!is.na(AchievementLevel))

ggplot(science_map_filtered) + 
geom_sf(aes(fill = Data)) +
  facet_wrap(~AchievementLevel) +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title = "Science Achievement (2015)", caption="Data from most recent year available") +
  theme_void()

```


```{r}

statebins(science_2015 %>% filter(AchievementLevel == "At or above basic"),
state_col = "NAME",
value_col = "Data") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Science Achievement, Basic (2019)") +
  theme_statebins("right")
```


```{r}
statebins(science_2015 %>% filter(AchievementLevel == "At or above proficient"),
state_col = "NAME",
value_col = "Data") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Science Achievement, Proficient (2019)") +
  theme_statebins("right")

```


```{r}
ggplot(science_2015, aes(state=NAME, fill=Data)) +
  geom_statebins() +
  coord_equal() +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  facet_wrap(~AchievementLevel) +
  labs(title="Science Achievement (2015)") +
  theme_statebins("right")
```


```{r}
statebins(math_2019 %>% filter(AchievementLevel == "At or above basic"),
state_col = "NAME",
value_col = "Data") +
    scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Math Achievement, Basic (2019)") +
  theme_statebins("right")
```


```{r}
statebins(math_2019 %>% filter(AchievementLevel == "At or above proficient"),
state_col = "NAME",
value_col = "Data") +
    scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Math Achievement, Proficient (2019)") +
  theme_statebins("right")
```


```{r}
ggplot(math_2019, aes(state=NAME, fill=Data)) +
  geom_statebins() +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  facet_wrap(~AchievementLevel) +
  labs(title="Math Achievement (2019)") +
  theme_statebins("right")
```

**Question 3**

a. How has funding for education changed over time? 

b. How does funding differ between states?

c. How have changes in funding differed between states?

```{r} 
#, fig.width=4, fig.height=2.5
funding_plot <- funding %>% 
  filter(Location=="United States") %>% 
  ggplot(aes(x=TimeFrame, y=Data)) +
  geom_line(aes(group=1)) +
  labs(x="Year", y = "Dollars", title = "Per-student spending (US average)", caption="Not adjusted for inflation")+
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  theme(axis.text.x = element_text(size=6, angle = 45))
```



```{r}
#For this question, I’m planning on looking at state-level data and achievement (both basic and proficient), as state-level educational expenditures (on a per-student basis) allows us to compare relationships between different states and funding types. I’m not sure whether I’ll look at this cross-sectionally, or longitudinally.

#As step 1 for this research question, I might include some plots just characterizing funding on a per-student basis between different states. 


#To show funding alone, I could make a circular bar plot (https://r-graph-gallery.com/circular-barplot.html). I’m thinking of using this because it will allow me to include data for all 50 states in one graph, without making the graph too overwhelming. I’ll either group regions together, or color-code by region. 


#Alternately, I might make a map visualization of funding, which would use color to show funding levels. To compare funding and achievement, I’ll make separate plots for each educational subject. 


# filter for just states
filtered_funding <- funding %>%  
  filter(LocationType=="State") %>% 
# average across years
  select(Location, TimeFrame, Data) %>% 
  group_by(Location) %>% 
  summarise(avg_funding = mean(Data, na.rm=T)) %>%  #, across(Region)
  ungroup() %>% 
  mutate(observation = 1:n())


# ----- This section prepare a dataframe for labels ---- #
# Get the name and the y position of each label
label_data <- filtered_funding
 
# calculate the ANGLE of the labels
number_of_bar <- nrow(label_data)
angle <-  90 - 360 * (label_data$observation-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
 
# calculate the alignment of labels: right or left
# If I am on the left part of the plot, my labels have currently an angle < -90
label_data$hjust<-ifelse( angle < -90, 1, 0)
 
# flip angle BY to make them readable
label_data$angle<-ifelse(angle < -90, angle+180, angle)
# ----- ------------------------------------------- ---- #
 
 
```



```{r}
#Plot

# Start the plot
p <- ggplot(filtered_funding, aes(x=as.factor(Location), y=avg_funding)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  
  # This add the bars with a blue color
  geom_bar(stat="identity", fill=alpha("skyblue", 0.7)) +
  
  # Limits of the plot = very important. The negative value controls the size of the inner circle, the positive one is useful to add size over each bar
  ylim(-100,18000) +
  
   labs(title = "Per-student expenditures", subtitle="Averaged over 1999-2019") +   #

  
  # Custom the theme: no axis title and no cartesian grid
  theme_minimal() +
  theme(
    #plot.title = element_text(hjust = 0.5, unit(c(2.1, 0, -1, 0), "cm")),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(.5,4), "cm")      # Adjust the margin to make in sort labels are not truncated!
  ) +
  
  # This makes the coordinate polar instead of cartesian.
  coord_polar(start = 0) +
  
  # Add the labels, using the label_data dataframe that we have created before
  geom_text(data=label_data, aes(x=observation, y=avg_funding+100, label=Location, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE )
 
p


# Add title, caption

#ggplotly(p)

```


```{r}
#Plot ordered by values
filtered_funding_ordered <- filtered_funding %>% 
  arrange(avg_funding) %>% 
  mutate(observation_ordered = 1:n())
# ----- This section prepare a dataframe for labels ---- #
# Get the name and the y position of each label
label_data_ordered <- filtered_funding_ordered
 
# calculate the ANGLE of the labels
number_of_bar_ordered <- nrow(label_data_ordered)
angle_ordered <-  90 - 360 * (label_data_ordered$observation_ordered-0.5) /number_of_bar_ordered     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
 
# calculate the alignment of labels: right or left
# If I am on the left part of the plot, my labels have currently an angle < -90
label_data_ordered$hjust<-ifelse( angle_ordered < -90, 1, 0)
 
# flip angle BY to make them readable
label_data_ordered$angle_ordered<-ifelse(angle_ordered < -90, angle_ordered+180, angle_ordered)
# ----- ------------------------------------------- ---- #


# Start the plot

p_ordered <- ggplot(filtered_funding_ordered, aes(x=as.factor(observation_ordered), y=avg_funding)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  
  # This add the bars with a blue color
  geom_bar(stat="identity", fill=alpha("skyblue", 0.7)) +
  
  # Limits of the plot = very important. The negative value controls the size of the inner circle, the positive one is useful to add size over each bar
  ylim(-100,19000) +
  
  labs(title = "Per-student expenditures", subtitle="Averaged over 1999-2019") +  
  
  # Custom the theme: no axis title and no cartesian grid
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(.25,4), "cm")      # Adjust the margin to make in sort labels are not truncated!
  ) +
  
  # This makes the coordinate polar instead of cartesian.
  coord_polar(start = 0) +
  
  # Add the labels, using the label_data dataframe that we have created before
  geom_text(data=label_data_ordered, aes(x=observation_ordered, y=avg_funding+100, label=Location, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data_ordered$angle_ordered, inherit.aes = FALSE ) 
 
p_ordered

```



```{r}
#Line Plot for state funding

funding %>% 
  filter(LocationType=="State") %>% 
  ggplot(aes(x=TimeFrame, y=Data, group = Location, color=Location)) +
  geom_line() +
    gghighlight(max(Data) > 20000|(max(Data) < 9250)) +
  scale_color_manual(values=met.brewer("Hiroshige", 10)) + #Derain
  labs(x="Year", y = "Dollars", title = "Per-student spending (US average)", caption="Not adjusted for inflation") +
  theme(axis.text.x = element_text(size=6, angle = 45, vjust = 1, hjust=1))


#highlight US average, maybe highest and lowest
 # gghighlight(Location==("United States")|(max(Data) > 20000)|(max(Data) < 92500)) +
```




```{r}
#Map with averaged funding

renamed_filtered_funding <- filtered_funding %>% 
  rename(NAME = Location) 

funding_map <- left_join(st, renamed_filtered_funding)

funding_map <- shift_geometry(funding_map,position = 'below')

funding_map_filtered <- funding_map %>% 
  filter(!is.na(avg_funding))

ggplot(funding_map_filtered) + 
geom_sf(aes(fill = avg_funding)) +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title = "Per-student expenditures", subtitle = "Averaged across 1999-2019") +
  theme_void()


#maybe add plotly

# try cartogram heatmap?
# statebins package

```


```{r}
statebins(filtered_funding,
state_col = "Location",
value_col = "avg_funding") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Per-student expenditures", subtitle="Averaged over 1999-2019") +
  theme_statebins("right")
```


```{r, eval = F}
#Funding vs Achievement, Faceted


# funding
#cut off at 2015


# science achievement
#use only above basic & above proficient

# math achievement
#use only above basic & above proficient

#stack them
grid.arrange(funding_plot, science_plot_basic, science_plot_proficient)

grid.arrange(funding_plot, math_plot_basic, math_plot_proficient)

grid.arrange(funding_plot, science_plot_basic, math_plot_basic)

plot_grid(funding_plot, science_plot_basic, math_plot_proficient, ncol=1, rel_widths = c(1, 1.5, 1.5))


#nested grids
bottom_row <- plot_grid(science_plot_basic, math_plot_basic, label_size = 12)

plot_grid(funding_plot, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1)
```



**Question 4**
How does funding track with educational achievement in math and science?

```{r}
# spending on top
# basic performance on both math and science underneath
plot_grid(funding_plot, subject_achievement_plot, ncol=1, rel_widths = c(1, 1.5))

```