---
title: "Final Project"
author: "Lucy Whitmore"
date: "3/13/2023"
output:
  ioslides_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

require(flexdashboard)
require(ggplot2)
require(plotly)
library(tidyverse)
library(gridExtra)
library(plotly)
library(cowplot)
library(tigris)
library(gghighlight)
library(statebins)
library(MetBrewer)
library(here)
```

```{r, echo=F, include=FALSE}
#Load data
math_achievement <- rio::import("~/Documents/EDLD-652/finalproject/Data/Eighth-grade math achievement levels.xlsx")
science_achievement_1 <-  rio::import("~/Documents/EDLD-652/finalproject/Data/Eighth-grade science achievement levels.xlsx")
science_achievement_2 <-  rio::import("~/Documents/EDLD-652/finalproject/Data/Eighth-grade science achievement levels (1).xlsx")
funding <-  rio::import("~/Documents/EDLD-652/finalproject/Data/Per-pupil educational expenditures.xlsx")
```


```{r, echo=F, include=FALSE}
#Rearrange data
# Math 
#Rename column and covert data to numeric
math_achievement <- math_achievement %>% 
  rename("AchievementLevel" = "Achievement Level") %>% 
  mutate(Data = as.numeric(Data), TimeFrame = as.numeric(TimeFrame)) %>% 
   mutate(AchievementType = case_when(grepl("basic", AchievementLevel) ~ "Basic",
                            grepl("proficient", AchievementLevel, ignore.case = TRUE) ~"Proficient"),
          Subject = "Math") 
# Science
#rbind both science achievement files to create one with all years, rename column, and convert data to numeric
science_achievement <- rbind(science_achievement_1, science_achievement_2) %>% 
  rename("AchievementLevel" = "Achievement Level") %>% 
  mutate(Data = as.numeric(Data), TimeFrame = as.numeric(TimeFrame)) %>% 
   mutate(AchievementType = case_when(grepl("basic", AchievementLevel) ~ "Basic",
                            grepl("proficient", AchievementLevel, ignore.case = TRUE) ~"Proficient"),
          Subject = "Science") 
# Funding
#Convert numbers from character to numeric
funding$Data <- as.numeric(funding$Data)

# Math, 2019
math_2019 <- math_achievement %>% 
         filter(LocationType=="State" & TimeFrame==2019 & (AchievementLevel == "At or above basic" | AchievementLevel=="At or above proficient")) %>% 
  rename(NAME = Location) 

#Science, 2015
science_2015 <- science_achievement %>% 
         filter(LocationType=="State" & TimeFrame==2015 & (AchievementLevel == "At or above basic" | AchievementLevel=="At or above proficient")) %>% 
  rename(NAME = Location) 
```


```{r, echo=F, include=FALSE}
#Set defaults
theme_set(theme_minimal())

options(ggrepel.max.overlaps = 25)
```
## Final Project

My research questions focused on educational achievement and educational funding. I looked at how both of those variables changed over time, and how they differed scross states within the US.

## Data

The data used for this project come from the KIDS COUNT Study, which is conducted by the Annie E. Casey Foundation (https://datacenter.kidscount.org/). Data is collected at both the national level and the state level within the US. The KIDS COUNT project contains data related to a number of different aspects of child development and well-being, including education, health, risky behavior, and family/community relationships. 

For my project, I looked at 8th grade achievement in math and science, as well as per-student educational spending. Data were available at the state and national level for all variables. Funding data was collected every year, and achievement data were collected about every two years.

## Visualization 1: Maps

Goal: Create maps showing state-level differences in educational achievement

Issues & Improvements:

- Loading times

- Unnecessary message printing

- Visibility w/ faceting

Solution:

- Statebins package

- Using dashboard to organize rather than faceting

## Map Plot

```{r}
statebins(science_2015 %>% 
  filter(AchievementLevel == "At or above basic"),
  state_col = "NAME",
  value_col = "Data") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title= "Science achievement has regional differences", subtitle = "Science Achievement, Basic (2015)", fill = "Proportion at or above basic") +
  theme_statebins("right")
```

## Map Plot Code

```{r, echo =T}
state_plot <- statebins(science_2015 %>% filter(AchievementLevel == "At or above basic"),
state_col = "NAME",
value_col = "Data") +
  scale_fill_gradientn(colors=met.brewer("Morgenstern")) +
  labs(title="Science Achievement, Basic (2015)") +
  theme_statebins("right")
```

## Visualization 2: State Funding

Goal: Show average per-student funding for all states in one readable plot.

Issues & Improvements:

- Readability

- Label alignment

- Sorting & Grouping 


## Original Version
```{r}
# filter for just states
filtered_funding <- funding %>%  
  filter(LocationType=="State") %>% 
# average across years
  select(Location, TimeFrame, Data) %>% 
  group_by(Location) %>% 
  summarise(avg_funding = mean(Data, na.rm=T)) %>%  #, across(Region)
  ungroup() %>% 
  mutate(observation = 1:n())


# ----- This section prepare a dataframe for labels ---- #
# Get the name and the y position of each label
label_data <- filtered_funding
 
# calculate the ANGLE of the labels
number_of_bar <- nrow(label_data)
angle <-  90 - 360 * (label_data$observation-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
 
# calculate the alignment of labels: right or left
# If I am on the left part of the plot, my labels have currently an angle < -90
label_data$hjust<-ifelse( angle < -90, 1, 0)
 
# flip angle BY to make them readable
label_data$angle<-ifelse(angle < -90, angle+180, angle)
# ----- ------------------------------------------- ---- #
```

```{r}
#Plot

# Start the plot
p <- ggplot(filtered_funding, aes(x=as.factor(Location), y=avg_funding)) +       # Note that id is a factor. If x is numeric, there is some space between the first bar
  
  # This add the bars with a blue color
  geom_bar(stat="identity", fill=alpha("skyblue", 0.7)) +
  
  # Limits of the plot = very important. The negative value controls the size of the inner circle, the positive one is useful to add size over each bar
  ylim(-100,18000) +
  
   labs(title = "Per-student expenditures", subtitle="Averaged over 1999-2019") +   #

  
  # Custom the theme: no axis title and no cartesian grid
  theme_minimal() +
  theme(
    #plot.title = element_text(hjust = 0.5, unit(c(2.1, 0, -1, 0), "cm")),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = unit(rep(.5,4), "cm")      # Adjust the margin to make in sort labels are not truncated!
  ) +
  
  # This makes the coordinate polar instead of cartesian.
  coord_polar(start = 0) +
  
  # Add the labels, using the label_data dataframe that we have created before
  geom_text(data=label_data, aes(x=observation, y=avg_funding+100, label=Location, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE )
 
p
```




```{r, include=F}
# filter for just states
filtered_funding <- funding %>%  
  filter(LocationType=="State") %>% 
# average across years
  select(Location, TimeFrame, Data) %>% 
  group_by(Location) %>% 
  summarise(avg_funding = mean(Data, na.rm=T)/100) %>%  #, across(Region)
  ungroup() %>% 
  #mutate(observation = 1:n()) %>% 
  mutate(quantile = as.factor(ntile(avg_funding, 4)))

# Set a number of 'empty bar' to add at the end of each group
empty_bar <- 4
to_add <- data.frame( matrix(NA, empty_bar*nlevels(filtered_funding$quantile), ncol(filtered_funding)) )
colnames(to_add) <- colnames(filtered_funding)
to_add$quantile <- rep(levels(filtered_funding$quantile), each=empty_bar)
funding_groups <- rbind(filtered_funding, to_add)
funding_groups <- funding_groups %>% arrange(quantile)
funding_groups$id <- seq(1, nrow(funding_groups))

#create labels
label_data <- funding_groups
number_of_bar <- nrow(label_data)
angle <- 90 - 360 * (label_data$id-0.5) /number_of_bar     # I substract 0.5 because the letter must have the angle of the center of the bars. Not extreme right(1) or extreme left (0)
label_data$hjust <- ifelse( angle < -90, 1, 0)
label_data$angle <- ifelse(angle < -90, angle+180, angle)

# prepare a data frame for base lines
base_data <- funding_groups %>% 
  group_by(quantile) %>% 
  summarize(start=min(id), end=max(id) - empty_bar) %>% 
  rowwise() %>% 
  mutate(title=mean(c(start, end)))

```


```{r, include=F}
#Plot
p <- ggplot(funding_groups, aes(x=as.factor(id), y=avg_funding, fill=quantile)) +      
  geom_bar(stat="identity", alpha=0.5) +
  # Limits of the plot = very important. The negative value controls the size of the inner circle, the positive one is useful to add size over each bar
  ylim(-100,185) +
   labs(title = "Educational spending varies across states", subtitle="Per-student expenditures averaged over 1999-2019") +   #
  theme_minimal() +
  theme(
    #plot.title = element_text(hjust = 0.5, unit(c(2.1, 0, -1, 0), "cm")),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.position="none",
    plot.margin = unit(rep(.5,4), "cm")      # Adjust the margin to make in sort labels are not truncated!
  ) +
  coord_polar(start = 0) +
  # Add the labels, using the label_data dataframe that we have created before
  geom_text(data=label_data, aes(x=id, y=avg_funding+10, label=Location, hjust=hjust), color="black", fontface="bold",alpha=0.6, size=2.5, angle= label_data$angle, inherit.aes = FALSE ) +
    geom_segment(data=base_data, aes(x = start, y = -5, xend = end, yend = -5), colour = "black", alpha=0.8, size=0.6 , inherit.aes = FALSE) +
  geom_text(data=base_data, aes(x = title, y = -18, label=quantile), hjust=c(1,1,0,0), colour = "black", alpha=0.8, size=4, fontface="bold", inherit.aes = FALSE)
```

## Improved Version
```{r} 
p
```

